---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# biomeUtils  

`biomeUtils` is part of the RIVM-ToolBox project aimed at providing standard set of tools that interact with open tools for a wide array of data analytics, including microbiomics. The RIVM-ToolBox is a set of individual R tools focused towards different goals/functionalities.  

* `biomeUtils` : Data handling  
    Outputs for standard data generating pipelines/workflows.  
  
* `biomeStats`: Data analytics  
    Common data analytics including basic statistics.  
  
* `biomeViz`: Data visualization  
    Data visualization of different data types.  

## Setup  

### Install  

```{r install, eval=FALSE}
devtools::install_github("RIVM-IIV-Microbiome/biomeUtils")
```

### Load pkg  
```{r setup}
library(biomeUtils)
```

## Data handling    

### Data  

```{r}
data("FuentesIliGutData")
FuentesIliGutData
```

### Get tibbles  

```{r}
data("FuentesIliGutData")
# get otu table in tibble format
otu_tib <- getAbundanceTibble(FuentesIliGutData,
                              select_rows = c("ASV302", "ASV82", "ASV410", "ASV2332"),
                              select_cols = c("sample_1", "sample_5", "sample_6"),
                              column_id = "FeatureID")

# get taxa table in tibble format
tax_tib <- getTaxaTibble(FuentesIliGutData,
                         select_rows = c("ASV302", "ASV82", "ASV410", "ASV2332"),
                         select_cols = c("Phylum", "Genus"),
                         column_id = "FeatureID")

# get sample data in tibble format
meta_tib <- getSampleTibble(FuentesIliGutData,
                            select_rows = c("sample_1", "sample_5", "sample_6"),
                            select_cols = c("participant_id", "ILI", "age"),
                            column_id = "FeatureID")
```

### Subset/Filter phyloseq  

```{r}

# Filter by phyloseq like subset_*
ps.filtered.samples  <- filterSampleData(FuentesIliGutData,
                                         ILI == "C" & BMI < 26)
ps.filtered.samples

ps.filtered.taxa  <- filterTaxaData(FuentesIliGutData,
                                    Phylum=="Firmicutes")

ps.filtered.taxa

# Filter by names like prune_*
sams.select <- sample_names(FuentesIliGutData)[1:10]
ps.filter.by.sam.names <- filterSampleByNames(FuentesIliGutData,
                                               ids = sams.select,
                                               keep = TRUE)

tax.select <- taxa_names(FuentesIliGutData)[1:10]
ps.filter.by.tax.names <- filterTaxaByNames(FuentesIliGutData,
                                           ids = tax.select,
                                           keep = TRUE)
```

### Check for polyphyly  

Check for polyphyletic taxa in `tax_table`. Useful to check this before aggregating at any level. Here, for e.g. _Eubacterium_ is is in both `Lachnospiraceae` and `EUbacteriaceae` family.    
```{r}
library(biomeUtils)
library(dplyr)
data("FuentesIliGutData")
polydf <- checkPolyphyletic(FuentesIliGutData,
                            taxa_level = "Genus",
                            return_df = TRUE)
polydf
```


### Compare phyloseqs  

```{r}
library(biomeUtils)
data("FuentesIliGutData")
ps1 <- filterSampleData(FuentesIliGutData, ILI == "C")

ps2 <- filterSampleData(FuentesIliGutData, ILI == "L1")

ps3 <- filterSampleData(FuentesIliGutData, ILI == "L2")


ps.list <- c("C" = ps1, "L1" = ps2, "L2" = ps3)

comparePhyloseq(ps.list)
```

## Calculations  

### Phylogenetic diversity  

```{r eval=FALSE}
library(biomeUtils)
data("FuentesIliGutData")
# reduce size for example
ps1 <- filterSampleData(FuentesIliGutData, ILI == "C")

meta_tib <- calculatePD(ps1, justDF=TRUE)
# check
meta_tib[c(1,2,3),c("PD", "SR")]
```

### Prevalence  

```{r}

data("FuentesIliGutData")
prev_tib <- getPrevalence(FuentesIliGutData,
                          return_rank= c("Family", "Genus"),
                          return_taxa = c("ASV4", "ASV17" , "ASV85", "ASV83"),
                          sort=TRUE)
head(prev_tib)
```

### Microbiota Uniqueness  
Extracts the minimum value from a \code{dissimilarity} matrix for each individual. This is the dissimilarity of an individual from their nearest neighbor. Here, the option of using a one or more reference samples is provided. see the man page. The original article (cite when using this) did all versus all samples dissimilarities. 
Wilmanski T, et al., (2021) Gut microbiome pattern reflects healthy ageing and predicts survival in humans. [Nature metabolism](https://www.nature.com/articles/s42255-021-00348-0)  

```{r fig.width=4,fig.height=3}
library(biomeUtils)
data("FuentesIliGutData")
ps <- getProportions(FuentesIliGutData)
dist.mat <- phyloseq::distance(ps, "bray")
muniq <- uniqueness(ps,
                    dist_mat=dist.mat,
                    reference_samples = NULL)
#head(muniq)

ggplot(muniq,aes(age,uniqueness)) +
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm") +
  theme_bw()

```

### Fold Difference   

```{r}
library(biomeUtils)

data("FuentesIliGutData")
# Keep only two groups
ps1 <- filterSampleData(FuentesIliGutData, ILI != "L2")

taxa_fd <- calculateTaxaFoldDifference(ps1, group="ILI")
# check
head(taxa_fd)
```

### Melting  
```{r}

library(biomeUtils)
data("FuentesIliGutData")
ps <- filterSampleData(FuentesIliGutData, ILI != "L2")
ps <- phyloseq::rarefy_even_depth(ps)
dist.mat <- phyloseq::distance(ps, "bray")
dist.melt.sample <- meltDistanceToTable(ps,
                                        dist_mat = dist.mat,
                                        name_dist_column = "Bray-Curtis",
                                        select_cols = c("participant_id", "ILI"))
head(dist.melt.sample)

```


# Pipe'ing' steps  

Read `%>%` as `and then`  
Below we use four R packages in combination to check for differences in ASVs between two groups. 

```{r fig.height=4, fig.width=6, eval=T}
library(biomeUtils)
library(ggplot2)
library(microbiome)
library(dplyr) # pipe function
data("FuentesIliGutData")

# Take FuentesIliGutData phyloseq object and then
FuentesIliGutData %>% 
  # join genus and species columns 
  uniteGenusSpeciesNames() %>% 
  # Remove 'L1' group and keep those with BMI less than 30 and then
  filterSampleData(ILI != "L1" & BMI < 30) %>% 
  # convert to rel. abundance and then
  microbiome::transform(., "compositional") %>% 
  # select ASVs with 0.0001 in at least 1% samples and then
  microbiome::core(., detection = 0.0001, prevalence = 0.01) %>% 
  # calculate foldchange and then 
  calculateTaxaFoldDifference(group="ILI")  %>% 
  # select major fold change ASVs and then plot 
  filter(abs(FoldDifference) >= 0.5 & 
           Prevalence.C != 0 &  # Remove ASVs not detected in Controls  
           Prevalence.L2 != 0) %>% # Remove ASVs not detected in L2
  # join genus name to FeatureID
  mutate(asv.genus = paste0(FeatureID, "-", Genus)) %>% 
  # Plot fold difference Notice that from here on below we use '+'
  ggplot(aes(FoldDifference, reorder(asv.genus, FoldDifference))) +
  geom_col(aes(fill=Enriched)) + 
  ylab("ASVs") + 
  scale_fill_manual(values = c( C = "steelblue", L2 = "brown3")) +
  theme_bw()

```



```{r}
sessionInfo()
```

