---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# biomeUtils  

`biomeUtils` is part of the RIVM-ToolBox project aimed at providing standard set of tools that interact with open tools for a wide array of data analytics, including microbiomics. The RIVM-ToolBox is a set of individual R tools focused towards different goals/functionalities.  

* Data handling  
    Outputs for standard data generating pipelines/workflows.  
  
* Data analytics  
    Common data analytics including basic statistics.  
  
* Data visualization  
    Data visualization of different data types.  

## Setup  

### Install  

```{r install, eval=FALSE}
devtools::install_github("microsud/biomUtils")
```

### Load pkg  
```{r setup}
library(biomeUtils)
```

## Data handling    

### Data  

```{r}
data("FuentesIliGutData")
FuentesIliGutData
```

### Get tibbles  

```{r}
data("FuentesIliGutData")
# get otu table in tibble format
otu_tib <- getAbundanceTibble(FuentesIliGutData,
                              select_rows = c("ASV302", "ASV82", "ASV410", "ASV2332"),
                              select_cols = c("sample_1", "sample_5", "sample_6"),
                              column_id = "FeatureID")

# get taxa table in tibble format
tax_tib <- getTaxaTibble(FuentesIliGutData,
                         select_rows = c("ASV302", "ASV82", "ASV410", "ASV2332"),
                         select_cols = c("Phylum", "Genus"),
                         column_id = "FeatureID")

# get sample data in tibble format
meta_tib <- getSampleTibble(FuentesIliGutData,
                            select_rows = c("sample_1", "sample_5", "sample_6"),
                            select_cols = c("participant_id", "ILI", "age"),
                            column_id = "FeatureID")
```

### Subset/Filter phyloseq  

```{r}

# Filter by phyloseq like subset_*
ps.filtered.samples  <- filterSampleData(FuentesIliGutData,
                                         ILI == "C" & BMI < 26)
ps.filtered.samples

ps.filtered.taxa  <- filterTaxaData(FuentesIliGutData,
                                    Phylum=="Firmicutes")

ps.filtered.taxa

# Filter by names like prune_*
sams.select <- sample_names(FuentesIliGutData)[1:10]
ps.filter.by.sam.names <- filterSampleByNames(FuentesIliGutData,
                                               ids = sams.select,
                                               keep = TRUE)

tax.select <- taxa_names(FuentesIliGutData)[1:10]
ps.filter.by.tax.names <- filterTaxaByNames(FuentesIliGutData,
                                           ids = tax.select,
                                           keep = TRUE)
```


### Compare phyloseqs  

```{r}
library(biomeUtils)
data("FuentesIliGutData")
ps1 <- filterSampleData(FuentesIliGutData, ILI == "C")

ps2 <- filterSampleData(FuentesIliGutData, ILI == "L1")

ps3 <- filterSampleData(FuentesIliGutData, ILI == "L2")


ps.list <- c("C" = ps1, "L1" = ps2, "L2" = ps3)

comparePhyloseq(ps.list)
```

## Calculations  

### Phylogenetic diversity  

```{r}
library(biomeUtils)
data("FuentesIliGutData")
# reduce size for example
ps1 <- filterSampleData(FuentesIliGutData, ILI == "C")

meta_tib <- calculatePD(ps1, justDF=TRUE)
# check
meta_tib[c(1,2,3),c("PD", "SR")]
```

### Prevalence  

```{r}

data("FuentesIliGutData")
prev_tib <- getPrevalence(FuentesIliGutData,
                          return_rank= c("Family", "Genus"),
                          return_taxa = c("ASV4", "ASV17" , "ASV85", "ASV83"),
                          sort=TRUE)
head(prev_tib)
```

### Fold Difference   

```{r}
library(biomeUtils)

data("FuentesIliGutData")
# Keep only two groups
ps1 <- filterSampleData(FuentesIliGutData, ILI != "L2")

taxa_fd <- calculateTaxaFoldDifference(ps1, group="ILI")
# check
head(taxa_fd)
```


```{r}
sessionInfo()
```

